Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-28 threshold/draft.Rnw
working on the examples in the paper

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
Department of Mathematics,\\University of Rhode Island
\and
Michael Kane\\
Yale University
\and
Bryan Lewis\\
Paradigm4, Inc.\\
\and
Alex Poliakov\\
Paradigm4, Inc.
}

@@

Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix $A$
(including possibly the case $i=j$). Denote the correlation of these two
vectors by $\mathrm{cor}(a_i,a_j) = a_j^T a_i$.  The following simple lemma
establishes a relationship between correlation and Euclidean distance between
two vectors.

@@

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors
below the correlation threshold
with respect to the ordering defined by $P$.
For example,
including only one more term finds that 49\% of adjacent vectors fall below the
a correlation threshold of $0.95$ in the example shown in Figure \ref{fig1} and
are pruned by identifying indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
leaving only 15 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

Algorithm \ref{prune} guarantees that no pruned vector pair exceeds the given
correlation threshold--the pruning does not produce false negatives.
The algorithm typically prunes the vast majority of pairs of vectors below the
correlation threshold with truncated singular value decompositions of only
relatively low rank $p$.  The choice of $p$ represents a balance between work in
computing a truncated SVD plus work required by step 4 and filtering the
thresholded values in step 7.

@@

@@

A full/naive thresholded correlation computation requires $O(n^2 m)$ flops. Let
$p$ be the selected IRLBA dimension. Then the proposed method requires
$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in $v_1$
that
meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
plus the flops required by IRLBA and the post-processing step of evaluating the
correlation of all the candidate pairs.

@@

\section{NOTES not for publication}

A full/naive thresholded correlation requires $O(n^2 m)$ flops. Let $p$ be the
selected IRLBA dimension. Then the proposed method requires $O(\ell  n  p)$
flops, where $\ell$ is the longest run of ordered entries in $v_1$ that meet
the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$ (not counting
from flops due to IRLBA and the post-processing step of evaluating the
correlation of all the candidate pairs).  One problem with the method as
written is that $\ell$ can a large fraction of n.

Unnecessarily large values of $\ell$ are due to the loose bound
$\sqrt{2(1-t)}/\sigma_1$ used in step 3 of Algorithm\ref{prune} which is only
sharp when $k=1$. General problems may not need to run through differencing
steps up to $\ell$ in step 4 of the algorithm. We can further improve the
efficiency of the algorithm by dynamically reducing $\ell$ in step 4 as
follows.

Equation \ref{vdist} expresses the Euclidean distance between any two columns
of the matrix $A$ in terms of singular components. Write the smallest squared
gap between entries of a particular singular vector $v_h$ as:
\[
\delta_h = \min_{i,j} (e_i^T v_h - e_j^T v_h)^2.
\]
Then note that
\begin{equation}\label{mindist}
\|a_i - a_j\|^2 \ge \sigma_1^2 \delta_1 + \sigma_2^2 \delta_2 + \cdots + \sigma_k^2\delta_k,
\end{equation}
for any indices $i,j$.

Fix $1\le \tilde{i}\le \ell$ and let $\nu \in R^p$ be the vector of minimum values in
each column of the $n\times p$ matrix $D^{\tilde{i}}PV_{1:p}$ from step 4 of
the algorithm.
It follows from Equation \ref{mindist} that
\[
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \ge \nu^Ts_{1:p}
\]
for all $\tilde{j}=1,2,\ldots,n$.

XXX need monotonicity of $\nu^Ts_{1:p}$ XXX

\section{The distribution of $||a_i - a_j||^2$}

\begin{lemma}
\label{asymdistm1lemma}
Let $m=1$, $X_i$ and $X_j$ be distributed as $\mathcal{N}(0, \Delta_{ij})$,
with $\Delta_{ii} = 1$.  Then
\begin{equation}
\label{eqasymdist1}
|| X_i - X_j ||^2 \sim R + \left(\Delta_{ij} + 1 \right) S + 
  \left(\Delta_{ij} - 1 \right) T
\end{equation}
where $R$ is distributed as $\chi_2^2$ and $S$ and $T$ are $\chi_1^2$.
\end{lemma}
\begin{proof}
\begin{equation*}
||X_i - X_j||^2 = X_i^2 - 2 X_i X_j + X_j^2
\end{equation*}
$X_i^2$ and $X_j^2$ are clearly $\chi_1^2$. To derive the distribution
of the cross-product term realize
\begin{equation*}
X_i X_j = \frac{1}{4} \left(X_i + X_j\right)^2 - 
  \frac{1}{4} \left(X_i - X_j\right)^2.
\end{equation*}
$X_i + X_j$ and $X_i - X_j$ are both distributed as normal and they are 
independent
since $\mathbf{E}X_1^2 = \mathbf{E}X_2^2$. Therefore, their squares
are $\chi_1^2$ and the leading coefficients in Equation \ref{eqasymdist1}
are the respective variances.
\end{proof}

\begin{lemma}
\label{asymdistlemma}
Let $a_i$ and $a_j$ be any two columns of the matrix 
$B \in \mathcal{R}^{m \times n}$, sampled from 
$\mathcal{N}(0, \Delta/m)$ where $\Delta_{ii} = 1$ for $1 \leq i \leq n$.
Then
\begin{equation*}
||a_i - a_j||^2 \xrightarrow[m]{D} \mathcal{N} \left(2 - 2 \Delta_{ij},  
  \frac{6 + 4\Delta_{ij}^2}{m} \right) 
\end{equation*}
\end{lemma}
This result follows by applying the central limit theorem to
the result in Lemma \ref{asymdistm1lemma} where $m$ is appropriately large.

\section{Conclusion}\label{conclusion}

@@

@@

@@

>>>>

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
University of Rhode Island
\and
Michael Kane\\
Yale University
\and
Bryan Lewis\\
Paradigm4, Inc.\\
\and
Alex Poliakov\\
Paradigm4, Inc.
}

@@

Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix $A$
(including possibly the case $i=j$). Denote the correlation of these two
vectors by $\mathrm{cor}(a_i,a_j) = a_j^T a_i$.  The following simple lemma
establishes a relationship between correlation and Euclidean distance between
two zero mean, unit norm vectors.

@@

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors below the correlation threshold with respect to the
ordering defined by $P$.  Including only one more term finds that 49\% of
adjacent vectors fall below the a correlation threshold of $0.95$ in the
example shown in Figure \ref{fig1} and are pruned by identifying indices such
that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
leaving only 15 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

Algorithm \ref{prune} guarantees that no pruned vector pair exceeds the given
correlation threshold--the pruning does not produce false negatives.  The
algorithm typically prunes the vast majority of pairs of vectors below the
correlation threshold with truncated singular value decompositions of only
relatively low rank $p$.  The choice of $p$ represents a balance between work
in computing a truncated SVD and evaluating $\ell$ rank $p$ matrix products in
Step 4 (each increasing in work as $p$ increases), and the work required to
filter the thresholded values in step 7 (decreasing work as $p$ increases).

@@

A full/naive thresholded correlation computation requires $O(n^2 m)$ flops.
Let $p$ be the selected IRLBA dimension. Then the proposed method requires
$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in
$v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
plus the flops required by IRLBA and the post-processing step 7 (also $O(np)$).

@@

We evaluted the algorithm using three publically-available gene expression
datasets.  The tests were performed on a desktop PC with a single AMD A10-7850K
3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running
the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R version
3.1.2 using double-precision arithmetic and the OpenBLAS library version
0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=4}.  All examples use the reference R implementation
function, {\tt tcor()}, which can be installed for the R language from the
development version of IRLBA GitHub repository using the {\tt devtools} package
with:

<<eval=FALSE,echo=TRUE>>=
devtools::install_github("bwlewis/irlba")
@

\subsection{BicatYeast small example}

The first example is a tiny proof of concept that provides the data for Figure
\ref{fig1} and illustrates our timing and memory-use measurement methodology.
Anyone can quickly run the expample to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
software\cite{biclust, bicat}. The data consist of 70 experiments invovling the
Saccharomyces Cerevisiae yeast measuring gene expression data across 419
Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
example finds all pairs of columns (Affymetrix probes) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
\begin{small}
<<eval=FALSE,echo=TRUE>>=
library(irlba)
library(biclust)  # for the BicatYeast data
data(BicatYeast)  # in gene by sample orientation
A = t(BicatYeast) # to correlate genes to genes
threshold = 0.95

# Compute using tcor():
m1  = sum(gc()[,2])  # Memory in megabytes currently in use
t1  = proc.time()
tx  = tcor(A, t=threshold, p=10)
t1  = (proc.time() - t1)[3] # Time in seconds
m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test

cat("tcor time (s)                      ", t1, "\n")
cat("tcor peak excess memory (MB)       ", m1, "\n")

# Compute using R's native cor() function:
m2  = sum(gc()[,2])
t2  = proc.time()
cx  = cor(A)                # full correlation matrix
diag(cx) = -Inf             # ignore diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2

cat("brute force time (s)               ", t2, "\n")
cat("brute force peak excess memory (MB)", m2, "\n")

# Compare results!
cat("brute force vector pairs found", nrow(pairs)/2, "\n")
cat("tcor vector pairs found       ", nrow(tx[[1]]), "\n")
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this tiny first example,
both in speed and slightly in memory use.  Each algorithm identifies the same
115 vector pairs that exceed the specified correlation threshold of 0.95, with
comparative timing and memory use shown in the following table.

\begin{center}
\begin{tabular}{l|rr}
 & {\tt tcor()} & native {\tt cor()} \\
Wall-clock time (s) & 0.26 & 0.02 \\
Excess peak memory use (MB) & 29.2 & 26.2 \\
\end{tabular}
\end{center}

@@

Brute-force computation benefits substantially from sustained use of BLAS Level
3 operations (matrix multiplication), achieving high utilization of available
CPU floating point capacity. The tcor method also benefits from optimized BLAS
routines, but not as much as it mostly makes use of Level 2 operations and
Level 3 operations on sequences of smaller problems.

%<<echo=FALSE, results=tex>>=
%require(xtable)
%print(
%    xtable(head(iris),
%        caption = "blah", label = "tab:five", table.placement = "!h", caption.placement = "top")
%  , size = "small", table.placement="ht")
%@

\section{Extensions}\label{extensions}
\section{Conclusion}\label{conclusion}

@@

%% BicatYeast data
\bibitem{bicat} Barkow, S., Bleuler, S., Prelic, A., Zimmermann, P., and E. Zitzler. BicAT: a biclustering analysis toolbox, Bioinformatics, 2006 22(10):1282-1283.

@@

%% R's biclust pacakge
\bibitem{biclust}
Kaiser, Sebastian, Santamaria, Rodrigo, Khamiakova, Tatsiana, Sill, Martin,
Theron, Roberto, Quintales, Luis, Leisch, Friederich, and De Troyer, Ewoud, 
\url{https://cran.r-project.org/web/packages/biclust}.

@@

\bibitem{tcor}
Lewis, Bryan,  \url{https://github.com/bwlewis/covar/blob/master/threshold/tcor.R}.
